Concurrent online sampling, for all, for free

### 2022-07-09 paper/main.tex
[wip] paper

@@

    \section{Solution}
        % What is online sampling
        Online sampling tries to address these issues by keeping the sample up to date while inserting new elements into the Database. Because every element that should be inserted is taken into consideration by the algorithm the sample is always up to date. Furthermore, no random I/O is required because no rows need to be retrieved from the disk.

        % Requirements

        % The algorithm it self

>>>>

% For images 
\usepackage{graphicx}
\usepackage{wrapfig}

\graphicspath{ {./images/} }

@@

    \section{Solution}
        % What is online sampling
        Online sampling tries to address these issues by keeping the sample up to date while inserting new tuples into the Database. Because every tuple that is going to be inserted is taken into consideration by the algorithm the sample is always up to date. Furthermore, no random I/O is required because no rows need to be retrieved from the disk. Modern DBMS use multiple threads for inserting therefore the algorithm also needs to support multithreading with very little overhead.
        % Requirements
        In consequence the requirements are:
        \begin{enumerate}
            \item Keep Samples up to date while inserting new tuples
            \item As little overhead as possible, per thread and overall
            \item Use constant amount of memory
            \item Easy integration into existing solutions
            \item Minimum amount of shared memory writes (as less locks as possible)
        \end{enumerate}

    \subsection{The skip node}
        % The idea behind skip length
        For every tuple that is inserted into the database the algorithm needs to determine whether the tuple should also be added to the sample or not. The tuples which are not inserted are "skipped" and tuples which are inserted into the sample are called reservoir tuples. Because the sampling strategy used by this algorithm is based on reservoir sampling [source], herby incoming tuples are inserted into the reservoir with some probability.

        \begin{wrapfigure}{r}{0.4\textwidth}
            \centering
            \includegraphics[height=5cm]{figure1.pdf}
            \caption{Example of two skips with different lengths}
        \end{wrapfigure}
        But deciding for every tuple separately using the probability method can be rather expensive. Therefore, the algorithm generates a length of a skip instead. The length is stored with some additional metadata in skip node and can be used to determine how many tuples should be skipped until the next reservoir tuple. A skip node that contains valid data is also called a skip. To determine whether a tuple is a reservoir tuple or not using the skip length is very cheap. First the skip length needs to be decremented by one and if the length reached zero the current tuple is a reservoir tuple and a new one skip needs to be generated. Figure one shows an example for two skips. The white tuples are skipped and the gray tuples are reservoir tuples.

        % Multithreaded skip nodes
        Hence, the skips are already stored in self contained nodes, these nodes can easily be distributed among multiple threads. Only the generation and distribution needs to be synchronized, 
        % Explain ownership
        once a thread owns a node no more synchronization is required and the thread can work independently. Every thread receives an independent stream of tuples that should be inserted and therefore also sampled. The amount of tuples is unknown and each thread could stop receiving tuples at any moment.

        % Preallocation
        To ensure the constant use of memory it is necessary that the maximum amount of threads that can be used for inserting is known when the algorithm is initialized. In consequence all skip nodes can be preallocated, one for every thread and one additional node as reserve (the reason for the reserve node is explained later on). Hence, a pointer to a skip node can simply be an index into the node array.

        % States of a skip node
        \begin{figure}[h]
            \includegraphics[height=3.5cm]{figure2.pdf}
            \caption{States of a skip node and the transitions between the states}
        \end{figure}
        One additional field that is stored in every skip node, is a successor pointer. This pointer can be used to arrange nodes in a singly linked list. There are three states a skip node can be in, initially all nodes are in the free list. The free list contains all nodes that are currently not used by any thread and nodes that do not contain valid data. If a skip should be generated a node needs to be allocated, thus it needs to be removed from the free list and the data of skip can be stored in this node. Afterwards the node is added to the list of skips (LOS). The LOS contains all nodes that store a valid skip, but are not currently owned by a thread. Additionally the LOS should always contain at least on skip. When a thread needs acquire a new skip, it can pop the first item of the LOS. Afterwards the thread might need to generate a new skip to ensure that the LOS is not empty. Figure 2 illustrates the lifecycle of a skip node.

    \subsection{Sampling while inserting}
